Large-scale spatiotemporal variation in vital rates and population dynamics of an alpine bird

### 2024-01-25 Manuscript_Nater_etal.qmd
Edited the abstract

Quantifying the temporal and spatial variation in animal demography is a central theme of ecology and important for directing policy. However, field sampling at large spatial extents for long periods of time is cost prohibitive and often politically untenable. Citizen science programs can alleviate these issues by harnessing stakeholders to increase the spatial and temporal resolution of datasets. Yet, most citizen science programs lack data from marked animals making it difficult to model demography. Integrated population models (IPM) offer promise to use multiple streams of data on populations or multiple aspects of a single dataset to estimate key vital rates for populations or communities. We leveraged a large-scale citizen science monitoring program---the "Hønsefuglportalen"--- to estimate spatial and temporal variation in demography for willow ptarmigan (*Lagopus lagopus*). The core field sampling approach was line-transect distance sampling with age-specific detections---the latter critical to modelling the population dynamics in the IPM. We explored latitudinal and longitudinal variation in recruitment, survival, and density; patterns of density dependence; and the relationship between recruitment and rodent occupancy using an IPM in a Bayesian framework following open science principles. Overall results were consistent with expectations of ecological theory. Slower life history patterns were observed at higher latitudes and altitudes portending differential effects of climate change on ptarmigan across their range. Observed patterns between rodents and recruitment were consistent with the alternative prey hypothesis (APH); however, spatial variation in the directionality and magnitude of effect were substantial. The robustness of the hierarchical structure of our model was instrumental in elucidating these patterns. Our study highlights the power of citizen science, IPMs, and open science principles to estimate spatial and temporal variation in the demography for a culturally important species.

>>>>

Quantifying temporal and spatial variation in animal population size and demography is a central theme in ecological research and important for directing management and policy. However, this requires field sampling at large spatial extents and over long periods of time, which is not only prohibitively costly but also often politically untenable. Citizen science programs can alleviate these constraints by recruiting stakeholders and the public to increase the spatial and temporal resolution of monitoring effort and hence resulting data. While the majority of citizen science programs are limited by opportunistic sampling designs, we are starting to see the emergence of structures citizen science programs that employ trained volunteers to collect data according to standardized protocols. Simultaneously, there is much ongoing development of statistical models that are increasingly more powerful and able to make more efficient use of field data. Integrated population models (IPMs), for example, are able to use multiple streams of data from different field monitoring programmes and/or multiple aspects of single datasets to estimate population sizes and key vital rates. Here, we developed a multi-area version of a recently developed integrated distance sampling model (IDSM) and applied it to data from a large-scale citizen science monitoring program---the "Hønsefuglportalen"--- to estimate spatial and temporal variation in population dynamics of willow ptarmigan (*Lagopus lagopus*) in Norway. We set up an open and reproducible workflow for exploring temporal, spatial (latitudinal, longitudinal, altitudinal), and residual variation in recruitment, survival, and population density, as well as relationships between vital rates and relevant covariates and signals of density dependence. Recruitment rates varied more across space than over time, while the opposite was the case for survival. Slower life history patterns (higher survival, lower recruitment) appeared to be more common at higher latitudes and altitudes, portending differential effects of climate change on ptarmigan across their range. While there was variation in the magnitude of the effect small rodent abundance had on recruitment, the relationships were predominantly positive and thus consistent with the alternative prey hypothesis, which predicts higher reproductive output when there is plenty of alternative prey for shared predators. Notably, the accurate estimation of covariate effect was only made possible by integrating data from several monitoring areas for analysis. Our study thus highlights the potential of citizen science monitoring and integrated modelling approaches for estimating and understanding spatio-temporal patterns species abundance and demographic rates, and showcases how corresponding workflows can be set up in a reproducible and semi-automated way that increases their usefulness for informing management and regular reporting towards national and international biodiversity frameworks.
